Type Checker Generation using Reference Attribute Grammars

### 2024-01-14 MSc.tex
Further iteration

  Static type checkers are a popular method for early detection of bugs in computer programs, by ensuring the correct data types are used throughout the program.
  One approach for implementing type checkers are Reference Attribute Grammars (RAGs), a formalism for specifying the static semantics of programming languages by declaring attributes on an abstract tree-representation of the program, and is used by the JastAdd meta-compilation system.

  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from syntax-directed typing rules.
  This aims to provide integration with existing compilers implemented with JastAdd, as well as a playground for type systems usable as an educational tool.

  We find RAGs to be a flexible model for syntax-directed typing rules, and our software was able to support simple values, nodes with premises placing demands on their children, and those with rules requiring type variables.
  It proved difficult to provide compile-time verification of the typing rules, and further work is required to extend the compiler to support type environments, a necessity for most real-world type systems.
}

@@

% Define the problem you are solving (or trying to solve).
%\CR{Prior work has already looked into pluggable type checkers.
%  Why is translating them into JastAdd interesting?
%  (aside from theoretical questions, you can also use
%  pragmatic reasons, such as integration into our courses.)}
In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of Reference Attribute Grammars (RAGs).
JastAdd is an interesting target because it has already been used to implement compilers of several languages, to which our typing rule implementations could be integrated.
JastAdd is also used extensively in courses at the department of computer science at LTH, where our project could serve as an educational tool, providing students an interactive way to define and test simple type systems.

Typing rules form a complex, often ambiguous, and infinite relation.
To be able to translate them into JastAdd code, we will need to define a limited subset of syntax-directed typing relations.
We identify a number of different kinds of typing rules, used in previous work in the field\ref{Pierce}, which share a common set of features for us to implement.
We then devise a strategy to translate these rules into snippets of iterative Java code, which can be evaluated by JastAdd.

@@

\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
  We evaluate the approach against Foo, and demonstrate Bar.
  Our case studies show Quux.
  Overall, this initial exploration shows \ldots''}

@@

\chapter{Background}
%\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinking ``I know why I want to read this''}
This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
Section \ref{abstractsyntaxtrees} explains abstract syntax trees, a simplified tree-based representation of a program, suitable for performing static analysis on.
These are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
%\CR{Consider saying just a little more about them here, for readers who have no idea what ASTs are and why they matter for
%    analysing programs.}
Section \ref{typetheory} details type systems,
%\CR{Too much detail in the following:}
by explaining each of the component pieces, followed by a description of their use within static type checking.
%\CR{You can use \texttt{\textbackslash label\{labelname\}} and \texttt{\textbackslash ref\{labelname\}} to reference (sub)secttions, that's usually clearer than ``next section''.}

@@

This makes it easier to write analysis based on the behaviour of a program, without concerning oneself with its exact formulation.
Several different formulations of a program with identical behaviour may be abstracted into identical ASTs.
The hierarchical structure of the tree lends itself especially well for recursive analyses, and eases the passing of information between connected structures by connecting them as parent and children nodes.
%\CR{what is ``this information'', concretely?}
A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if-else node, with named branches pointing towards the condition and the two values.
%\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
%\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}

\CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}
\begin{figure}[h]
\begin{lstlisting}[]
Expr   -> Expr "+" Term | Term
Term   -> Term "*" Factor | Factor
Factor -> Int | "(" Expr ")"
\end{lstlisting}
  \caption{\reviewquestion{One like this? I'm unsure how to tie it in effectively.}}
  \label{typesyntaxspec}
\end{figure}

@@

\subsection{Data types}
%\CR{think about the connection between types and values here.  You may even want to talk about values in a separate subsection.}
Data types can be considered a categorisation of values which share common properties, such as supporting numerical operations or accessing a certain element in a list of characters.
%several different types of data in one program, such as numbers or text.
Most programming languages come with a number of built-in data types, from different kinds of numbers (integers, floating point) to strings or lists and arrays.

@@

\CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
  For each concrete (and possibly very complex) operator on the left-hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
  This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractions over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
}

@@

\subsubsection{Type syntax}
A type syntax, or set of types, is a definition of the valid types within a language.
\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
\reviewquestion{I believe Pierce simply refers to it as the "set of types". I seem to have got the term type syntax from EDAP15 \url{https://fileadmin.cs.lth.se/cs/Education/EDAP15/2022/web/skills/type-system.html}. In my opinion it works better to have a clear name for it, so I'm keeping it for now, unless you have further thoughts.}
It can consist of fundamental types, such as integers or booleans.
There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
This enables differentiating lists that store integers from those that store strings, and ensuring that any item added to a list must be of the same type as the type parameter.
%\CR{A bit vague; consider giving an example.}
\begin{figure}[h]
\begin{lstlisting}[]
abstract Type;

@@

\subsection{Non-Terminal Attributes}
\cite{higherorderattributegrammars}
\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are
not a necessary feature for RAGs.  Please introduce them in a separate section: you can very directly connect them to the idea of ``type syntax''.}

@@

\section{Language design}\label{sectionlanguage}
\begin{figure}[]
\begin{lstlisting}[language=java]
ruleset = rule_list

rule_list =
    rule
  | rule_list rule

rule =
    '[' upperid ']' premises_list '---+' formula
  | '[' upperid ']' formula

premises_list =
    formula
  | premises_list ',' formula

formula = term ':' tyterm

term =
    upperid '(' term_list ')'
  | upperid '(' ')'
  | upperid
  | lowerid

tyterm =
    upperid                                              // TyVal
  | lowerid                                              // TyVar

term_list =
    term
  | term_list ',' term
\end{lstlisting}
  \caption{Syntax definition for the typing rule language}
  \label{trastspec}
\end{figure}
\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
  This would be before the current 3.1, then.
  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
  of the object language as well as of the types.}

@@

\begin{figure}[]
  \resizebox{\textwidth}{!}{%
  \begin{tikzpicture}[
      every node/.style = {draw=black, align=center},
      every path/.style = {draw, -latex},
      artifact/.style={draw=black},
      process/.style={ellipse}
      %greennode/.style={shape=circle, draw=green, line width=2},
      %rednode/.style={shape=circle, draw=red, line width=2}
    ]
    \node (typerules)                               {Typing rules\\definition (\ref{typingruledefinition})};
    \node (trparsing) [right=of typerules, process] {Scan,\\parse,\\build};
    \node (trparser)  [above=of trparsing]          {Typing rule parser\\(\ref{typingrulesparser})};
    \node (trast)     [right=of trparsing]          {Typing rules AST\\(\ref{typingrulesast})};

    %\node (typecheckgen) [below=of trast, process, text width=5cm] {
    %  \textbf{Type checker generation}\\
    %  \hrule
    %  \begin{itemize}[noitemsep,topsep=2pt]
    %    \item Copy template
    %    \item Generate type checker from typing rules
    %  \end{itemize}
    %};
    \node (typecheckgen)       [below=of trast, process] {Generate\\type checker};
    \node (typingrulecompiler) [right=of trast]          {Typing rule\\compiler (\ref{sectionimplementation})};
    \node (typechecker)        [below=of typecheckgen]   {Type checker \\(\ref{typecheckingalgorithm})};

    \node (olast)     [below=of typerules] {AST specification\\(\ref{astdef})};

    \node (outputtypechecking) [below=of typechecker, process] {Perform\\typechecking};
    \node (inputast)           [left=of outputtypechecking]    {Input program\\AST};
    \node (outputparsing)      [left=of inputast, process]     {Scan,\\parse,\\build};
    \node (typeerrors)         [right=of outputtypechecking]   {Type errors\\(\ref{errorhandling})};
    \node (inputprog)          [left=of outputparsing]         {Input program};

    %\node (template) [below=of inputast, text width=6cm] {
    %  \textbf{Type checker template}\\
    %  \hrule
    %  \begin{itemize}[noitemsep,topsep=2pt]
    %    \item Scanner, parser, builder
    %    \item Project files
    %    \item Type checking algorithm (\ref{typecheckingalgorithm})
    %  \end{itemize}
    %};
    \node (templateparser)        [below=of outputparsing]                               {Tree-based\\parser\\(\ref{treebasedparser})};
    \node (templateprojectfiles)  at (inputast |- templateparser)                        {Project files\\(\ref{projectfiles})};
    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser)              {Type checking\\aspect (\ref{typecheckingaspect})};
    \node (templatetypesyntax)    [right=of typecheckingalgorithm]                       {Type syntax\\AST};
    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm) (templatetypesyntax)] {};
    \node                         [above, draw=none] at (template.north)                 {Support library};

    \node (userinput)   [dashed, fill=none, fit=(typerules) (olast)]                  {};
    \node               [above, draw=none] at (userinput.north)                       {Object language\\static semantics};
    %\node (typechecker) [dashed, fill=none, fit=(outputparsing) (outputtypechecking)] {};
    %\node               [above, draw=none] at (typechecker.north)                     {Type checker};

@@

    \matrix [below left, nodes={minimum size = 15pt, align=left}, align=left] at (current bounding box.north east) {
      \node [artifact, label=right:Artifact] {}; \\
      \node [process, label=right:Process] {}; \\
      \node [draw=none, label=right:Flow] (leg1){}; \\
      \node [draw=none, label=right:Flow (Temporary)] (leg2){}; \\
    };
    \draw [label=right:Line] (leg1.west) to (leg1.east);\\
    \draw [dashed, label=right:Dashed] (leg2.west) to (leg2.east);\\

  \end{tikzpicture}
  }
  \caption{An overview of the project components and flow.
It represents the flow of the program, with components and artefacts (represented as rectangular boxes) with arrows linking them to actions (elliptical boxes) which in turn have arrows to the new components and artefacts they produce.
The dashed arrows from represent links from the support library components that exist in the current experimental state of the project, but in a finished product would instead link to component from the object language's own compilation environment.
  \CR{Can you mark, e.g. via colours, the parts that the user provides,
           and the parts that will be in the generated output project?}}
  \label{overviewgraph}
\end{figure}

@@

%\subsection{Type checker generation}
The type checker generation is based on a single-responsibility principle, where each node of the abstract syntax tree (of \emph{which} of the three (or four?) ASTs?) is responsible for generating the code it corresponds to \CR{Not sure what you mean}.
Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.

@@

%\subsubsection{Aspect generation}
\subsection{Aspect generation}
The \lstinline{RuleSet} node is responsible for generating the overall structure of the output file, a JastAdd aspect called \lstinline{TypingRules}.
It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation.

%\subsubsection{Method implementation overview}
\subsection{Method implementation overview}
The implementation of a typing rule can range from a single return statement to a system of declarations of variables, type variables and control flow.
These are explained in further detail in subsequent sections, but here follows a quick overview.

Firstly, the algorithm checks if the number of premises is zero, in which case it immediately generates the conclusion, consisting of a return statement.
If there are premises, it first generates code for the declarations of variables and type variables.
This is followed by an if-statement to test whether the premises hold true, containing a return statement.
Should the premises not hold, a type error is produced.
\CR{What if there are multiple rules that apply to the same AST node?  E.g., one rule that allows \ty{Int} addition and one that allows \ty{String} addition.}

%\subsubsection{Parameters}
\subsection{Parameters}
When the typing rules concern a node with parameters, these can be named arbitrarily \CR{Who does the naming?  Why is the ``arbitrarily'' bit important?}\CR{ Will this work if the type system specification has a type variable called ``this'' or ``null'' or ``class''?}, and a variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
These variables are then utilised in later premises or type variable declarations.

%\subsubsection{Premises/control flow}
\subsection{Premises/control flow}
The rule's premises are translated into an if-statement, which checks all types using a method \lstinline{matches} added to all Type nodes.

The current implementation of \lstinline{Type.matches(Type t)} checks that the Type it is called on is either of the same class or a subtype of the type \lstinline{t}.
\CR{is ``matches'' a good name for something that is not symmetric?}
This allows the language limited sub-typing at the type checker generation step, though the pros and cons of this have not been evaluated.
\CR{Can you give an example of a case where this is helpful, and one example where it might be troublesome?
Let's use the ``if'' rule as an example.  If $\ty{Number} :> \ty{Int}$ (\ty{Number} is a supertype of \ty{Int}), what type will I infer if $t_2 = \ty{Number}$ and $t_3 = \ty{Int}$?  How about the other way around?}
It is also possible to override the \lstinline{matches} implementation for a specific Type, by adding a definition to a JastAdd aspect included in the output type checker.
\CR{This sounds like this is a way to support ``ad-hoc type polymorphism''?}

%\subsubsection{Type variables}
\subsection{Type variables}
In the case of type variables, the generated code takes a dynamic approach \CR{Meaning?} to simplify the implementation.
Each type variable that appears in the conclusion is instantiated to a null value, and the code generated by each premise first checks if the value is still null and either assigns the variable its own type or tests to make sure the existing type matches its own.
\CR{Is there a hidden order dependency? (Cf. my earlier question about ``matches'')}

\CR{Overall, I am not convinced that this strategy for presenting the type checking algorithm is very effective (though it might be better for someone who is reading the source code at the same time).
  I would propose to merge 3.2.2, 3.2.3, 3.2.4, 3.2.5, and to split them into (a) the type checking algorithm (while ignoring the code generation bit), and (b) the code generation aspect.
  I.e., for (a), pretend that you are interpreting the typing rules instead of compiling them, and please try to present an algorithm that captures
  the check.
  The algorithm can be in plain English with bullet points / numbered sub-steps, or in some pseudo-formal notation (``foreach premise $p$ in rule $r$ ...''), as long as it is clear.}

>>>>

  Static type checkers are a popular method for early detection of bugs in computer programs, by ensuring the correct data types are used throughout the program.
  % RAGs may not be the most convenient format for typing rules
  One approach for implementing type checkers are Reference Attribute Grammars (RAGs), a formalism for specifying the static semantics of programming languages by declaring attributes on an abstract tree-representation of the program, and is used by the JastAdd meta-compilation system.

  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from a subset of possible typing rules.
  This aims to provide integration with existing compilers implemented with JastAdd, as well as a playground for type systems usable as an educational tool.

  % 'Syntax-directed' needs explaining if used, "a subset known as syntax-directed"
  We find RAGs to be a flexible model for this subset of typing rules,
  % This is too specific, works with small languages from research literature
  and our software was able to support simple values, nodes with premises placing demands on their children, and those with rules requiring type variables.
  It proved difficult to provide compile-time verification of the typing rules, and further work is required to extend the compiler to support type environments, a necessity for most real-world type systems.
}

@@

% Define the problem you are solving (or trying to solve).
%\CR{Prior work has already looked into pluggable type checkers.
%  Why is translating them into JastAdd interesting?
%  (aside from theoretical questions, you can also use
%  pragmatic reasons, such as integration into our courses.)}
In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of Reference Attribute Grammars (RAGs).

Pacak et al. have previously implemented a compiler for generating incremental type checkers using declarative logic programming language Datalog\cite{Pacak}.
It focused on defining a way of expressing typing rules as a series of relations solvable by the Datalog compiler.
Our use of RAGs instead allows us to build our analysis directly onto a representation of the program's structure.
JastAdd is also an interesting target because it has already been used to implement compilers of several languages, to which our typing rule implementations could potentially be integrated.
It is also used extensively in courses within the Department of Computer Science at LTH, where our project could serve as an educational tool, providing students an interactive way to define and test simple type systems.

Typing rules form a complex, often ambiguous, and infinite relation.
To be able to translate them into JastAdd code, we will need to define a limited subset of syntax-directed typing relations.
We identify a number of different kinds of typing rules, used in previous work in the field\cite{Pierce}, which share a common set of features for us to implement.
We then devise a strategy to translate these rules into snippets of iterative Java code, which can be evaluated by JastAdd.

@@

We define a custom parsable language for typing rules adapted from natural semantics, and introduce a strategy to compile it into RAG attributes.
Our case studies show that simple type systems can be systematically compiled into attributes on the AST, and the tree syntax provides a natural approach for recursive evaluation.
Overall, this initial exploration shows that RAGs are a flexible approach for implementing type checkers for syntax-directed typing rules, though there are difficulties when implementing rule verification at compiler runtime.
%\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
%  We evaluate the approach against Foo, and demonstrate Bar.
%  Our case studies show Quux.
%  Overall, this initial exploration shows \ldots''}

@@

\chapter{Background}
%\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
%\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinking ``I know why I want to read this''}
This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
Section \ref{abstractsyntaxtrees} explains abstract syntax trees, a simplified tree-based representation of a program, suitable for performing static analysis on.
These are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
%\CR{Consider saying just a little more about them here, for readers who have no idea what ASTs are and why they matter for
%    analysing programs.}
Section \ref{typetheory} details type systems, by explaining each of the component pieces, followed by a description of their use within static type checking.

@@

This makes it easier to write analysis based on the behaviour of a program, without concerning oneself with its exact formulation.
Several different formulations of a program with identical behaviour may be abstracted into identical ASTs.
The hierarchical structure of the tree lends itself especially well for recursive analyses, and eases the passing of information between connected structures by connecting them as parent and children nodes.
%\CR{what is ``this information'', concretely?}
A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if node, with named branches pointing towards the condition and the two values.
%\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
%\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}

%\CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}
%\begin{figure}[h]
%\begin{lstlisting}[]
%Expr   -> Expr "+" Term | Term
%Term   -> Term "*" Factor | Factor
%Factor -> Int | "(" Expr ")"
%\end{lstlisting}
%  \caption{\reviewquestion{One like this? I'm unsure how to tie it in effectively.}}
%  \label{typesyntaxspec}
%\end{figure}

@@

\subsection{Data types}
%\CR{think about the connection between types and values here.  You may even want to talk about values in a separate subsection.}
Data types can be considered a categorisation of values which share common properties, such as supporting numerical operations or accessing a certain element in a list of characters.
While expressions may be arbitrarily complex, nested structures, by assigning it a type, we can perform useful analysis on it, without having to evaluate its value.
%\CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
%  For each concrete (and possibly very complex) operator on the left-hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
%  This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
%  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractions over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
%}

%several different types of data in one program, such as numbers or text.
Most programming languages come with a number of built-in data types, from different kinds of numbers (integers, floating point) to strings or lists and arrays.

@@

@@

\subsubsection{Type syntax}\label{typesyntax}
We're using the term `type syntax' to mean a definition of the valid types within a language, often referred to simply as the set of types.
%\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
%\reviewquestion{I believe Pierce simply refers to it as the "set of types". I seem to have got the term type syntax from EDAP15 \url{https://fileadmin.cs.lth.se/cs/Education/EDAP15/2022/web/skills/type-system.html}. In my opinion it works better to have a clear name for it, so I'm keeping it for now, unless you have further thoughts.}
It can consist of fundamental types, such as integers or booleans.
There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
This enables differentiating lists that store integers from those that store strings, and ensuring that any item added to a list must be of the same type as the type parameter.
%\CR{A bit vague; consider giving an example.}
\begin{figure}[h]
\begin{lstlisting}[]
abstract Type;

@@

%\subsection{Non-Terminal Attributes}
%Our project also uses not\cite{higherorderattributegrammars}
%\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are not a necessary feature for RAGs.
%Please introduce them in a separate section: you can very directly connect them to the idea of ``type syntax''.}

@@

\section{Language design}\label{sectionlanguage}
\begin{figure}[]
\begin{lstlisting}[]
ruleset = rule_list;

rule_list =
    rule
  | rule_list rule;

rule =
    '[' uppercase_id ']' premises_list '---+' formula
  | '[' uppercase_id ']' formula;

premises_list =
    formula
  | premises_list ',' formula;

formula = term ':' tyterm;

term =
    uppercase_id '(' term_list ')'
  | uppercase_id '(' ')'
  | uppercase_id
  | lowercase_id;

tyterm =
    uppercase_id
  | lowercase_id;

term_list =
    term
  | term_list ',' term;

uppercase_id = "[A-Z][A-Za-z0-9-]";
uppercase_id = "[a-z][A-Za-z0-9-]";
\end{lstlisting}
  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to BNF.}
  \label{trastspec}
\end{figure}
%\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
%  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
%  This would be before the current 3.1, then.
%  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
%  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
%  of the object language as well as of the types.}

@@

\begin{figure}[]
  \resizebox{\textwidth}{!}{%
  \begin{tikzpicture}[
      every node/.style = {align=center, line width=0.8pt},
      every path/.style = {draw, -latex},
      artifact/.style={draw=black},
      userinput/.style={draw=red},
      generated/.style={draw=blue},
      process/.style={ellipse, draw=black}
      %greennode/.style={shape=circle, draw=green, line width=2},
      %rednode/.style={shape=circle, draw=red, line width=2}
    ]
    \node (typerules) [userinput]                    {Typing rules\\definition (\ref{typingruledefinition})};
    \node (trparsing) [right=of typerules, process] {Scan,\\parse,\\build};
    \node (trparser)  [above=of trparsing, artifact]          {Typing rule parser\\(\ref{typingrulesparser})};
    \node (trast)     [right=of trparsing, generated]          {Typing rules AST\\(\ref{typingrulesast})};

    %\node (typecheckgen) [below=of trast, process, text width=5cm] {
    %  \textbf{Type checker generation}\\
    %  \hrule
    %  \begin{itemize}[noitemsep,topsep=2pt]
    %    \item Copy template
    %    \item Generate type checker from typing rules
    %  \end{itemize}
    %};
    \node (typecheckgen)       [below=of trast, process] {Generate\\type checker};
    \node (typingrulecompiler) [right=of trast, artifact]          {Typing rule\\compiler (\ref{sectionimplementation})};
    \node (typechecker)        [below=of typecheckgen, generated]   {Type checker \\(\ref{typecheckingalgorithm})};

    \node (olast)     [below=of typerules, userinput] {AST specification\\(\ref{astdef})};

    \node (outputtypechecking) [below=of typechecker, process] {Perform\\typechecking};
    \node (inputast)           [left=of outputtypechecking, generated]    {Input program\\AST};
    \node (outputparsing)      [left=of inputast, process]     {Scan,\\parse,\\build};
    \node (typeerrors)         [right=of outputtypechecking, generated]   {Type errors\\(\ref{errorhandling})};
    \node (inputprog)          [left=of outputparsing, artifact]         {Input program};

    %\node (template) [below=of inputast, text width=6cm] {
    %  \textbf{Type checker template}\\
    %  \hrule
    %  \begin{itemize}[noitemsep,topsep=2pt]
    %    \item Scanner, parser, builder
    %    \item Project files
    %    \item Type checking algorithm (\ref{typecheckingalgorithm})
    %  \end{itemize}
    %};
    \node (templateparser)        [below=of outputparsing, artifact]                               {Tree-based\\parser\\(\ref{treebasedparser})};
    \node (templateprojectfiles)  at (inputast |- templateparser)           [artifact]           {Project files\\(\ref{projectfiles})};
    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser) [artifact]           {Type checking\\aspect (\ref{typecheckingaspect})};
    \node (templatetypesyntax)    [right=of typecheckingalgorithm, artifact]                       {Type syntax\\AST (\ref{typesyntax})};
    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm) (templatetypesyntax), artifact] {};
    \node                         [above, draw=none] at (template.north)                 {Support library};

    \node (userinput)   [dashed, fill=none, fit=(typerules) (olast), artifact]                  {};
    \node               [above, draw=none] at (userinput.north)                       {Object language\\static semantics};
    %\node (typechecker) [dashed, fill=none, fit=(outputparsing) (outputtypechecking)] {};
    %\node               [above, draw=none] at (typechecker.north)                     {Type checker};

@@

    \matrix [left, nodes={minimum size = 15pt, align=left}, align=left, draw=black] at (current bounding box.north east) {
      \node [artifact, label=right:Artifact] {}; \\
      \node [userinput, label=right:User provided] {}; \\
      \node [generated, label=right:Generated] {}; \\
      \node [process, label=right:Process] {}; \\
      \node [draw=none, label=right:Flow] (leg1){}; \\
      \node [draw=none, label=right:Flow (Temporary)] (leg2){}; \\
    };
    \draw [label=right:Line] (leg1.west) to (leg1.east);\\
    \draw [dashed, label=right:Dashed] (leg2.west) to (leg2.east);\\

  \end{tikzpicture}
  }
  \caption{An overview of the project components and flow.
It represents the flow of the program, with components and artefacts (represented as rectangular boxes) with arrows linking them to actions (elliptical boxes) which in turn have arrows to the new components and artefacts they produce.
The dashed arrows from represent links from the support library components that exist in the current experimental state of the project, but in a finished product would instead link to component from the object language's own compilation environment.
  %\CR{Can you mark, e.g. via colours, the parts that the user provides, and the parts that will be in the generated output project?}
  }
  \label{overviewgraph}
\end{figure}

@@

%\subsection{Type checker generation}
%The type checker generation is based on a single-responsibility principle, where each node of the typing rule AST is responsible for generating the code it corresponds to \CR{Not sure what you mean}.
In this section we present our approach to translating typing rules.
We start by presenting pseudocode for our typing rule evaluation.
%Section \ref{typecheckingalgorithm} presenting a p the algorithm we use to evaluate typing rules in pseudocode.
This is followed by Section \ref{codegeneration}, which explains how we generate  code to implement this algorithm.

%For each rule:
%\begin{enumerate}
%  \item For each parameter in the conclusion:
%    \begin{enumerate}
%      \item Declare a variable by the same name, initialised by fetching the child of the same index.
%    \end{enumerate}
%  \item For each type variable used in the rule:
%    \begin{enumerate}
%      \item Initialise it to null
%      \item Iterate through each premise it appears on the right-hand side of:
%        \begin{enumerate}
%          \item If the type variable is set to null, evaluate the type of the left-hand side and assign it to the type variable.
%          \item If the type variable is not set to null, evaluate the type of our left-hand side and check if it is the same type or a subtype to the type variable. If it is not, throw a type error.
%        \end{enumerate}
%    \end{enumerate}
%  \item Check whether each of the premises is true:
%    \begin{enumerate}
%      \item Evaluate the type of the left-hand side of the premise and check if it is the same type or a subtype to the right-hand side.
%    \end{enumerate}
%    If all true: Return the right-hand side of the premise
%  \item Throw a type error if we reach this point
%\end{enumerate}

\begin{figure}[]
%foreach parameter in rule.conclusion.parameters:
%  declare var "parameer.name" = getChild(parameter.index)
\begin{lstlisting}[]
foreach typevar in rule.typevars:
  typevar.type = null;
  foreach premise in rule.premises where premise.rhs == typevar:
    if typevar.type == null:
      typevar.type = premise.lhs.type()
    else if typevar.type != premise.lhs.type():
      throw typeerror("Type variable mismatch")

// Check if all evaluated left-hand side types match their right-hand side
if allTrue(rule.premises.map(p => p.lhs.type() == p.rhs)):
  return rule.conclusion.rhs
else:
  throw typeerror("Typechecking failed")
\end{lstlisting}
  \caption{Pseudocode for the implementation of the type() attribute}
  \label{psuedotype}
\end{figure}

Here we aim to give an overview of the algorithm used to evaluate the typing rules.
Figure \ref{psuedotype} contains a pseudocode algorithm for the evaluation of a typing rule.
It is not representative of the actual generated code, but captures its behaviour in the manner of an interpreted language.

\subsection{Code generation}\label{codegeneration}

In this section we will present the code we generate to implement the previously discussed algorithm.
Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.
%TODO: Rewrite this section to be less bad

@@

\subsubsection{Aspect generation}
The \lstinline{RuleSet} node is responsible for generating the overall structure of the output file, a JastAdd aspect called \lstinline{TypingRules}.
It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation (though this feature is not yet fully working).

\subsubsection{Method implementation overview}
The implementation of a typing rule can range from a single return statement to a system of declarations of variables, type variables and control flow.
These are explained in further detail in subsequent sections, but here follows a quick overview.

Firstly, the algorithm checks if the number of premises is zero, in which case it immediately generates the conclusion, consisting of a return statement.
If there are premises, it first generates code for the declarations of variables and both declarations and unification for type variables.
This is followed by an if-statement to test whether the premises hold true, containing a return statement.
Should the premises not hold, a type error is produced.
%\CR{What if there are multiple rules that apply to the same AST node?  E.g., one rule that allows \ty{Int} addition and one that allows \ty{String} addition.}

\subsubsection{Parameters}
When the typing rules concern a node with parameters, these can be named arbitrarily in the typing rule.
%\CR{Who does the naming?  Why is the ``arbitrarily'' bit important?}
%\CR{ Will this work if the type system specification has a type variable called ``this'' or ``null'' or ``class''?}.
A variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
This naming format means any errors reported will use the same variable names as the user has defined in the typing rules, in order to minimise the disconnect when analysing errors from the generated code.
However, it also means it is not possible to name a parameter any of the reserved keywords in Java.
This could be alleviated by adding a prefix to the variable names, at the cost of a slight indirection for the user.
These variables are then utilised in later premises or type variable declarations.

\subsubsection{Type variables}
In the case of type variables, much of the logic is handled at runtime, to simplify the code generation.
Each type variable that appears in the conclusion is instantiated to a null value, and the code generated by each premise first checks if the value is still null and either assigns the variable its own type or tests to make sure the existing type matches its own.

The generated code could be simplified in future work by instantiating the type variable to the type of the left-hand side of the first premise it appears in, and then only checking if each of the subsequent uses match it.

\subsubsection{Premises/control flow}
For rules containing premises, the return statement is protected by an if-statement.
At this point we've declared variable for each of the child nodes and type variables used, so the premises can very directly be translated into boolean expressions.
A premise \lstinline{x : Bool} translates into the expression \lstinline{x.type().matches(new Bool());}.
This includes a call to the \lstinline{type()} of each left-hand side of the premise, recursively evaluating the necessary types.
Since our typing rules can only reference their direct children, this recursion will always progress down the program's AST, ensuring termination.
Each of the boolean expressions then joined by the \lstinline{&&} operator into a single if-statement, protecting the return of the right-hand side of the conclusion.
For the case where any of the boolean expressions is false, the if-statement is followed by a thrown exception representing a type error.

%\CR{Overall, I am not convinced that this strategy for presenting the type checking algorithm is very effective (though it might be better for someone who is reading the source code at the same time).
%  I would propose to merge 3.2.2, 3.2.3, 3.2.4, 3.2.5, and to split them into (a) the type checking algorithm (while ignoring the code generation bit), and (b) the code generation aspect.
%  I.e., for (a), pretend that you are interpreting the typing rules instead of compiling them, and please try to present an algorithm that captures
%  the check.
%  The algorithm can be in plain English with bullet points / numbered sub-steps, or in some pseudo-formal notation (``foreach premise $p$ in rule $r$ ...''), as long as it is clear.}
